---
output: 
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    number_sections: true
    toc: true
    toc_depth: 4
title: "SURVIVAL ANALYSIS - TCGA PRAD CANCER"
author: 
- Kelvin Ofori-Minta
- University of Texas at El Paso (UTEP) 
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontsize: 11pt
spacing: single
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb, bm}
- \usepackage{amsfonts}
- \usepackage{amsthm}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \rhead{Collaborative Research}
- \lhead{TCGA - Clinical Data}
- \cfoot{\thepage}
- \usepackage{algorithm}
- \usepackage[noend]{algpseudocode}
---

```{r setup, include=FALSE}
lung<-read.csv("lung_data2.csv") #predictors of mortality

# In Survival Analysis, you have three options for modeling the survival function: non-parametric (such as Kaplan-Meier), semi-parametric (Cox regression), and parametric (such as the Weibull distribution).Nov 25, 2018
```

```{r}
lung$Chemo<-factor(lung$Chemo, 
                     levels=c("No Chemo", "Chemo"),
                     labels=c("No Chemo", "Chemo"))

lung$AGE_cat<-factor(lung$AGE_cat, 
                     levels = c("50-60", "60-70", "70-80", "Above 80", "Below 50"),
                     labels = c("50-60", "60-70", "70-80", "Above 80", "Below 50"))

lung$SEX<-factor(lung$SEX,
                   levels=c("Female", "Male"),
                   labels=c("Female", "Male"))

lung$CDCC_TOTAL_BEST<-factor(lung$CDCC_TOTAL_BEST,
                             levels = c("0","1","2","3"),
                             labels = c("0","1","2","3"))

lung$TUMOR_SIZE_cat<-factor(lung$TUMOR_SIZE_cat,
                            levels=c("<=1cm","1cm-2cm","2cm-3cm","3cm-4cm","4cm-5cm"),
                            labels=c("<=1cm","1cm-2cm","2cm-3cm","3cm-4cm","4cm-5cm"))

lung$GRADE<-factor(lung$GRADE,
                   levels=c("Moderately differentiated","Poorly differentiated", "Undifferentiated","Unknown","Well differentiated"),
                   labels=c("Moderately differentiated","Poorly differentiated", "Undifferentiated","Unknown", "Well differentiated" ))

lung$Pathology<-factor(lung$Pathology,
                       levels = c("Adenocarcinoma","Other", "Squamous"),
                       labels = c("Adenocarcinoma","Other", "Squamous"))


lung$Visceral_Pleural_Invasion<-factor(lung$Visceral_Pleural_Invasion,
                                       levels = c("Other", "Present"),
                                       labels = c("Other", "Present"))

lung$LYMPH_VASCULAR_INVASION2<-factor(lung$LYMPH_VASCULAR_INVASION2,
                                      levels=c("Absent", "Present", "Unknown"),
                                      labels=c("Absent", "Present", "Unknown"))

lung$Margins<-factor(lung$Margins,
                     levels = c("Other","Positive","Zero"),
                     labels = c("Other","Positive","Zero"))

lung$Lymph_Nodes_Sampled<-factor(lung$Lymph_Nodes_Sampled,
                                 levels = c("<10",">=10", "Unknown"),
                                 labels = c("<10",">=10", "Unknown"))

lung$Excision_less_than1<-factor(lung$Excision_less_than1,
                                 levels = c("FALSE","TRUE"),
                                 labels = c("FALSE", "TRUE"))
```

\newpage
# Partition Data
```{r, warning=FALSE, message=FALSE}
require(caTools)
set.seed(1)
split = sample.split(lung$DX_LASTCONTACT_DEATH_MONTHS,SplitRatio = 0.85)
train=subset(lung, split==T)
test=subset(lung, split==F)
```


# Survival curve and logrank test for treatment groups
```{r, warning=FALSE, message=FALSE}
library("survival")
library("survminer")

fit1<-survfit(Surv(lung$DX_LASTCONTACT_DEATH_MONTHS,
                   lung$PUF_VITAL_STATUS) ~ lung$Chemo, data=lung)

#print(fit1)

summary(fit1)$table


d=data.frame(time=fit1$time, 
             n.risk=fit1$n.risk,
             n.event=fit1$n.event,
             n.censor=fit1$n.censor,
             surv=fit1$surv)
head(d)
```
\newpage
```{r}
ggsurvplot(fit1,
          #legend.labs=c("tumor_free", "with_tumor"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))
```
Logrank test is insignificant = no significant difference between the two treatment groups, but the curve shows a slightly higher survival probability for chemo groups than no_chemo groups. This is violation of ph assumption.

\newpage
# cumulative hazard plot of treatment groups
```{r}
ggsurvplot(fit1,
          #legend.labs=c("tumor_free", "with_tumor"),
          pval = TRUE, conf.int = F,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          #surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"),
          fun = "cumhaz")

```

\newpage
# Predicting survival probabability from survival object.
```{r, warning=FALSE}

library(peperr)
w=as.matrix(test[,-c(1,2)]) #matrix of covariates

e=predictProb(fit1, Surv(lung$DX_LASTCONTACT_DEATH_MONTHS,
                         lung$PUF_VITAL_STATUS), x=w,
              times=5)
#e #predicted probabilities of all of survival with chemo treatment at the 5th month
```



# CPH model for predictors of mortality 
```{r, warning=FALSE, message=FALSE}
library(rms)
lung$Chemo=relevel(as.factor(lung$Chemo), ref="No Chemo")
cph_fit = cph(Surv(DX_LASTCONTACT_DEATH_MONTHS,PUF_VITAL_STATUS) ~ Chemo +
                      AGE_cat+       
                      SEX +                     
                      CDCC_TOTAL_BEST  +         
                      TUMOR_SIZE_cat  +          
                      GRADE  +                 
                      Visceral_Pleural_Invasion+ 
                      LYMPH_VASCULAR_INVASION2+  
                      Margins  +                
                      Lymph_Nodes_Sampled +    
                      Excision_less_than1, 
                    data = lung, surv = TRUE, x=TRUE, y=TRUE)
#summary(cph_fit)
cph_fit$coefficients
exp(cph_fit$coefficients)
lp=exp(cph_fit$linear.predictors) #linear predictors for all patients

# plot(lung$DX_LASTCONTACT_DEATH_MONTHS, exp(cph_fit$linear.predictors),
#      type="p", main = "Hazard rates of all patients vs time", 
#      ylab = "exp(beta)", xlab = "time")

```
\newpage
# COXPH model for predictors of mortality 
```{r, include = FALSE}
lung$Chemo=relevel(as.factor(lung$Chemo), ref="No Chemo")
cox_fit1 <- coxph(Surv(DX_LASTCONTACT_DEATH_MONTHS,PUF_VITAL_STATUS) ~ Chemo +
                      AGE_cat+       
                      SEX +                     
                      CDCC_TOTAL_BEST  +         
                      TUMOR_SIZE_cat  +          
                      GRADE  +                 
                      Visceral_Pleural_Invasion+ 
                      LYMPH_VASCULAR_INVASION2+  
                      Margins  +                
                      Lymph_Nodes_Sampled +    
                      Excision_less_than1, 
                    data = lung)
summary(cox_fit1)
cox_fit1$coefficients #odds

exp(cox_fit1$coefficients) #HR


# plot(lung$DX_LASTCONTACT_DEATH_MONTHS, exp(cox_fit1$linear.predictors),
#      type="p", main = "Hazard rates of all patients vs time", 
#      ylab = "exp(beta)", xlab = "time")
```

\newpage
# format results of cox model
```{r}
require(kableExtra)
broom::tidy(cox_fit1 , 
            exp=TRUE) %>%
  kable()
```

\newpage
```{r, warning=FALSE, message=FALSE}
cox_fit1 %>%
  gtsummary::tbl_regression(exp=TRUE)
```



\newpage
# predict survival probability from cox model
```{r}
library(pec)
#quantiles of survival times
ttt=quantile(lung$DX_LASTCONTACT_DEATH_MONTHS)
#for selected predictor variables
testdata=test[,-c(1,2,4,10)]
#predicting at each quantile, one column for each quantile, one row for each testing individual 
pred=predictSurvProb(cph_fit, newdata=testdata, times=ttt) 
```

\newpage
# cumulative Hazard vs Survival Probability
```{r}
surv100=survest(cph_fit)

plot(surv100$surv,surv100$cumhaz, 
     main = "Plot of Cum. Hazard vs Survival Probability",
     xlab="survival", ylab = "Cum. Hazard")

```


# Smoothing of the Hazard Function - Non Parametric Survival
```{r}
# library(bshazard)
# mod2<-bshazard(Surv(lung$DX_LASTCONTACT_DEATH_MONTHS,
#                    lung$PUF_VITAL_STATUS) ~ 1, data=lung)
# 
# plot(mod2)
```



\newpage
# predicting risks
```{r}
train$Chemo=relevel(as.factor(train$Chemo), ref="No Chemo")
cox_fit2 <- coxph(Surv(DX_LASTCONTACT_DEATH_MONTHS,PUF_VITAL_STATUS) ~Chemo +
                      AGE_cat+       
                      SEX +                     
                      CDCC_TOTAL_BEST  +         
                      TUMOR_SIZE_cat  +          
                      GRADE  +                 
                      Visceral_Pleural_Invasion+ 
                      LYMPH_VASCULAR_INVASION2+  
                      Margins  +                
                      Lymph_Nodes_Sampled +    
                      Excision_less_than1, 
                    data = train)
#summary(cox_fit1)
cox_fit2 %>%
  gtsummary::tbl_regression(exp=TRUE)
```
\newpage
# Predict 
```{r}
lp1=predict(cox_fit2, newdata=train, type="lp")#linear predictor ("lp") for each 8826 subjects

#Train_Data
riskscore1_train=predict(cox_fit2, newdata=train, type="risk")#the risk score exp(lp)
hist(riskscore1_train)

# Test_Data
riskscore1_test=predict(cox_fit2, newdata=test, type="risk") #the risk score exp(lp)
hist(riskscore1_test, xlim = c(0,2))

#Entire Data
riskscore1_all=predict(cox_fit2, newdata=lung, type="risk")#the risk score exp(lp)
hist(riskscore1_all, xlim = c(0,2))

#Density plot of scores
# TRAIN_RISK <- data.frame(rs=riskscore1_train)
# TEST_RISK <- data.frame(rs=riskscore1_test)
ALL_DATA <- data.frame(rs=riskscore1_all)

TRAIN_RISK$type<-'train'
TEST_RISK$type<-'test'
ALL_DATA$type<-'entire_data'

ggplot(TEST_RISK, aes(rs, fill=type)) + geom_density(alpha = 0.2) 
ggplot(TRAIN_RISK, aes(rs, fill=type)) + geom_density(alpha = 0.2) 

datlen<-rbind(TRAIN_RISK,TEST_RISK,ALL_DATA)
ggplot(datlen, aes(rs, fill=type)) + geom_density(alpha = 0.2)
```

\newpage
```{r}
#obtain hazard ratio for train data
train$Chemo=relevel(as.factor(train$Chemo), ref="No Chemo")
cox_fit3 <- coxph(Surv(DX_LASTCONTACT_DEATH_MONTHS,PUF_VITAL_STATUS) ~ Chemo, 
                  data = train)
cox_fit3 %>%
  gtsummary::tbl_regression(exp=TRUE)

hr3=exp(cox_fit2$coefficients) #HR for chemo therapy
lphr3=predict(cox_fit2, train, type="lp") #predicted hazard ratio
plot(riskscore1_train, lphr3)


range(lphr3)
range(1-lphr3)
```

\newpage
```{r}
plot(riskscore1_train, lphr3, ylab = "Hazard Ratio" ,
     xlab="risk score", col="blue")

abline(h=0, lty=2, lwd=2, col="red")
abline(v=1, lwd=3, col="red")
abline(v=v[2], lwd=3, col="snow3")
abline(v=v[1], lwd=3, col="snow3")
abline(v=v[3], lwd=3, col="snow3")
abline(v=v[4], lwd=3, col="snow3")
```
***Comment***\n
A positive HR indicates worse conditions/prognosis, while a negative coefficient indicates a better condition/prognosis.
Riskscores > 1 shows increased hazards of mortality with multiple HRF.


\newpage
***Subtracting HR from 1 gives new scale of HR reads, where lower HR indicates worse conditions, and vice versa***
```{r}
plot(riskscore1_train, 1-lphr3, ylab = "Hazard Ratio" ,
     xlab="Risk Score", col="blue")# ylim=c(0.6,1.6), xlim=c(0.6,1.6)

abline(h=1, lty=2, lwd=2, col="red")
abline(v=1, lwd=3, col="red")
abline(v=v[2], lwd=3, col="snow3")
abline(v=v[1], lwd=3, col="snow3")
abline(v=v[3], lwd=3, col="snow3")
abline(v=v[4], lwd=3, col="snow3")
```


\newpage
```{r}
plot(riskscore1_train, 1-lphr3, ylab = "Hazard Ratio",
     xlab="risk score", col="blue", ylim=c(0.6,2))# ylim=c(0.6,1.6), xlim=c(0.6,1.6)
#abline(v=quantile(riskscore1_train), lwd=3, col="red")
abline(h=1, lty=2, lwd=2, col="red")
#abline(v=riskscore1_train==1,lwd=3, col="green")
abline(v=v[2], lwd=3, col="red")
abline(v=v[1], lwd=3, col="snow3")
abline(v=v[3], lwd=3, col="snow3")
abline(v=v[4], lwd=3, col="snow3")
```

When risk score is > 1.3, a survival benefit from chemotherapy can be expected 
as the risk score increases (i.e., multiple HRF present), survival
decreases and the potential benefit from AC increases



\newpage
\section{The scatter plot of hazard ratio of chemo(reference:no chemo)versus risk score}
```{r}
#obtain hazard ratio for train data
lung$Chemo=relevel(as.factor(lung$Chemo), ref="No Chemo")
cox_fitchemo <-coxph(Surv(DX_LASTCONTACT_DEATH_MONTHS,PUF_VITAL_STATUS) ~ Chemo, 
                  data = lung)
cox_fitchemo %>%
  gtsummary::tbl_regression(exp=TRUE)

hr_chemo=exp(cox_fitchemo$coefficients) #HR for chemo therapy
hr_chemo1=predict(cox_fitchemo, type="lp") #predicted hazard ratio
plot(riskscore1_all, 1-hr_chemo1)

range(hr_chemo1)
range(hr_chemo)
range(1-hr_chemo1)
```

\newpage
```{r}
plot(riskscore1_all, hr_chemo1, ylab = "Hazard Ratio" ,
     xlab="risk score", col="blue")

abline(h=0, lty=2, lwd=2, col="red")
abline(v=1, lwd=3, col="red")
abline(v=v[2], lwd=3, col="snow3")
abline(v=v[1], lwd=3, col="snow3")
abline(v=v[3], lwd=3, col="snow3")
abline(v=v[4], lwd=3, col="snow3")
```


\newpage
# Est.PH {survC1} - Derivation of a risk score by a Cox proportioal hazard model
## Obtain Risk scores from the best predictors of mortality
```{r}
#Provides risk score by fitting data to a Cox's proportional hazards model with a given set of predictors
# Input data. The 1st column should be time-to-event, and the 2nd column is event indicator (1=event, 0=censor). The rest of the columns are covariates/predictors used in the model. No character variable or missing is allowed. 
#OUTPUT
# beta = Estimates for regression coefficient in the Cox model
# var = Variance-Covariance matrix for the beta above
# rs 	= Risk score of each individual
# ft 	= coxph object with the fitted model

library(survC1)
train1=train[,c(1:2)] #time & status

train2 =train[, c(3:15)] # other covariates

#convert other sub levels in all categorical covariates to integer
p = data.frame(lapply(train2, as.integer))

#combine numeric time & status with the numeric covariates
train_data = data.frame(cbind(train1,p)) 

#Make sure distribution of variables are not distorted after conversion
require(inspectdf)
show_plot(inspect_cat(train)) # inspect categorical columns 

show_plot(inspect_num(train)) #inspect numeric columns 

show_plot(inspect_num(train_data)) #inspect numeric columns 
```

```{r}
#obtain risk scores for each individual
rsmodel=Est.PH(train_data)
riskscores=rsmodel$rs
hist(riskscores)
hist(sqrt(riskscores))
#riskscores1 = 1 - riskscores
```


```{r}
#obtain hazard rates
coef=rsmodel$beta
exp(coef)
hist(sqrt(riskscores))
#riskscores1 = 1 - riskscores
```


```{r}
lphr3=predict(cox_fit3, train, type="lp") #predicted hazard ratio
hrrr=1-lphr3
```



```{r}
#baseline cumulative hazard function for all covariates set to zero(centered=False)
bh=basehaz(cph_fit, centered=F)

#plot baseline hazard vs time
plot(bh$time, bh$hazard)
#length(bh$hazard) #945
```




# predict linear predictors for test data 
```{r, include=FALSE}
#lp=linear predictors, betas x covariates for each observation
lps=predict(cph_fit, test, type = "lp")
length(lps) #1500

#h_r= bh$hazard*(exp(lps)) #hazard_ratio
#hr=exp(lps)

#times of interest
time_of_interest=test$DX_LASTCONTACT_DEATH_MONTHS


#calculate the risk of event from the baseline hazard and linear predictors which returns a vector of probabilities for each person in the data.
sur=exp(-bh[time_of_interest,1])^exp(lps)
ri=1-sur;ri
#event_prob_t1=1-exp(-bh[time_of_interest,1])^exp(lps)
```



# cumulative incidence plot
<!-- cumulative incidence, also called incidence proportion, in epidemiology, estimate of the risk that an individual will experience an event or develop a disease during a specified period of time. Cumulative incidence is calculated as the number of new events or cases of disease divided by the total number of individuals in the population at risk for a specific time interval. Researchers can use cumulative incidence to predict risk of a disease or event over short or long periods of time. -->

<!-- McNutt, Louise-Anne and Krug, Allison. "cumulative incidence". Encyclopedia Britannica, Invalid Date, https://www.britannica.com/science/cumulative-incidence. Accessed 28 June 2022.  -->

```{r}
library(cmprsk)
s=cuminc(lung$DX_LASTCONTACT_DEATH_MONTHS, lung$PUF_VITAL_STATUS, cencode=1)
plot(s)
e=as.matrix(s)
```














