---
output: 
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    number_sections: true
    toc: true
    toc_depth: 4
title: "SURVIVAL ANALYSIS - NCDB"
author: 
- Kelvin Ofori-Minta
- University of Texas at El Paso (UTEP) 
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontsize: 11pt
spacing: single
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb, bm}
- \usepackage{amsfonts}
- \usepackage{amsthm}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \rhead{Summer RA}
- \lhead{NCDB Research}
- \cfoot{\thepage}
- \usepackage{algorithm}
- \usepackage[noend]{algpseudocode}
---
\newpage
\section{Loading Data and Preparations}
```{r setup, include=FALSE}
lung<-read.csv("lung_data2.csv") #predictors of mortality

```

```{r, include=FALSE}
lung$Chemo<-factor(lung$Chemo, 
                     levels=c("No Chemo", "Chemo"),
                     labels=c("No Chemo", "Chemo"))

lung$AGE_cat<-factor(lung$AGE_cat, 
                     levels = c("50-60", "60-70", "70-80", "Above 80", "Below 50"),
                     labels = c("50-60", "60-70", "70-80", "Above 80", "Below 50"))

lung$SEX<-factor(lung$SEX,
                   levels=c("Female", "Male"),
                   labels=c("Female", "Male"))

lung$CDCC_TOTAL_BEST<-factor(lung$CDCC_TOTAL_BEST,
                             levels = c("0","1","2","3"),
                             labels = c("0","1","2","3"))

lung$TUMOR_SIZE_cat<-factor(lung$TUMOR_SIZE_cat,
                            levels=c("<=1cm","1cm-2cm","2cm-3cm","3cm-4cm","4cm-5cm"),
                            labels=c("<=1cm","1cm-2cm","2cm-3cm","3cm-4cm","4cm-5cm"))

lung$GRADE<-factor(lung$GRADE,
                   levels=c("Moderately differentiated","Poorly differentiated", "Undifferentiated","Unknown","Well differentiated"),
                   labels=c("Moderately differentiated","Poorly differentiated", "Undifferentiated","Unknown", "Well differentiated" ))

lung$Pathology<-factor(lung$Pathology,
                       levels = c("Adenocarcinoma","Other", "Squamous"),
                       labels = c("Adenocarcinoma","Other", "Squamous"))


lung$Visceral_Pleural_Invasion<-factor(lung$Visceral_Pleural_Invasion,
                                       levels = c("Other", "Present"),
                                       labels = c("Other", "Present"))

lung$LYMPH_VASCULAR_INVASION2<-factor(lung$LYMPH_VASCULAR_INVASION2,
                                      levels=c("Absent", "Present", "Unknown"),
                                      labels=c("Absent", "Present", "Unknown"))

lung$Margins<-factor(lung$Margins,
                     levels = c("Other","Positive","Zero"),
                     labels = c("Other","Positive","Zero"))

lung$Lymph_Nodes_Sampled<-factor(lung$Lymph_Nodes_Sampled,
                                 levels = c("<10",">=10", "Unknown"),
                                 labels = c("<10",">=10", "Unknown"))

lung$Excision_less_than1<-factor(lung$Excision_less_than1,
                                 levels = c("FALSE","TRUE"),
                                 labels = c("FALSE", "TRUE"))
```

\newpage
\subsection{Partition Data}
```{r, warning=FALSE, message=FALSE}
require(caTools)
set.seed(1)
split = sample.split(lung$DX_LASTCONTACT_DEATH_MONTHS,SplitRatio = 0.85)
train=subset(lung, split==T)
test=subset(lung, split==F)
```

\newpage
\section{COXPH model for predictors of mortality - ALL DATA} 
```{r, warning=FALSE, message=FALSE}
library("survival")
# library("survminer")

lung$Chemo=relevel(as.factor(lung$Chemo), ref="No Chemo")
cox_fit1 <- coxph(Surv(DX_LASTCONTACT_DEATH_MONTHS,PUF_VITAL_STATUS) ~ Chemo +
                      AGE_cat+       
                      SEX +                     
                      CDCC_TOTAL_BEST  +         
                      TUMOR_SIZE_cat  +          
                      GRADE  +                 
                      Visceral_Pleural_Invasion+ 
                      LYMPH_VASCULAR_INVASION2+  
                      Margins  +                
                      Lymph_Nodes_Sampled +    
                      Excision_less_than1, 
                    data = lung)

cox_fit1$coefficients #odds

exp(cox_fit1$coefficients) #HR

# lung$Chemo=relevel(as.factor(lung$Chemo), ref="No Chemo")
# cox_fit11 <- coxph(Surv(DX_LASTCONTACT_DEATH_MONTHS,PUF_VITAL_STATUS) ~
#                       AGE_cat+       
#                       SEX +                     
#                       CDCC_TOTAL_BEST  +         
#                       TUMOR_SIZE_cat  +          
#                       GRADE  +                 
#                       Visceral_Pleural_Invasion+ 
#                       LYMPH_VASCULAR_INVASION2+  
#                       Margins  +                
#                       Lymph_Nodes_Sampled +    
#                       Excision_less_than1, 
#                     data = lung)
# 
# cox_fit11$coefficients #odds
# 
# exp(cox_fit11$coefficients) #HR

```

\newpage
# format results of cox model
```{r, warning=FALSE, message=FALSE}
require(kableExtra)
broom::tidy(cox_fit1 , 
            exp=TRUE) %>%
  kable()
```

\newpage
```{r, warning=FALSE, message=FALSE}
cox_fit1 %>%
  gtsummary::tbl_regression(exp=TRUE)
```


\newpage
\section{Predicting Risks scores and Hazard Ratio from COX PH Model}
\subsection{Distribution of Riskscores}
```{r, warning=FALSE, message=FALSE}
require(ggplot2)
riskscore1_all=predict(cox_fit1,  type="risk") #the risk score exp(lp)
hist(riskscore1_all, xlim = c(0,2), xlab = "Riskscores")

#Density plot of riskscores
# TRAIN_RISK <- data.frame(rs=riskscore1_train)
# TEST_RISK <- data.frame(rs=riskscore1_test)
ALL_DATA <- data.frame(Riskscores=riskscore1_all)

# TRAIN_RISK$type<-'train'
# TEST_RISK$type<-'test'
ALL_DATA$data<-'Full Data'
ggplot(ALL_DATA, aes(Riskscores, fill=data)) + geom_density(alpha = 0.5) 

# ggplot(TEST_RISK, aes(rs, fill=type)) + geom_density(alpha = 0.2) 
# ggplot(TRAIN_RISK, aes(rs, fill=type)) + geom_density(alpha = 0.2) 
# 
# datlen<-rbind(TRAIN_RISK,TEST_RISK,ALL_DATA)
# ggplot(datlen, aes(rs, fill=type)) + geom_density(alpha = 0.2)
```

\newpage
\section{Hazard Ratios}
```{r, warning=FALSE, message=FALSE}
lphr3=predict(cox_fit1, type="lp") #predicted hazard ratio

# hist(lphr3, xlim = c(0,2), xlab = "HR")
# hist(1-lphr3, xlim = c(0,2), xlab = "HR")
range(lphr3)
range(1-lphr3)


ALLDATA <- data.frame(Hazard_Ratios=lphr3)
ALLDATA$data<-'Full Data'
ggplot(ALLDATA, aes(Hazard_Ratios, fill=data)) + geom_density(alpha = 0.5) 

ALLDATA <- data.frame(Hazard_Ratios1=1-lphr3)
ALLDATA$data<-'Full Data'
ggplot(ALLDATA, aes(Hazard_Ratios1, fill=data)) + geom_density(alpha = 0.5)
```

\newpage
\section{Scatterplots of Riskscores vs Hazard Ratios}
```{r, warning=FALSE, message=FALSE}
plot(riskscore1_all, lphr3, ylab = "Hazard Ratio" ,
     xlab="risk score", col="blue")
v=quantile(riskscore1_all)
abline(h=0, lty=2, lwd=2, col="red")
abline(v=1, lwd=3, col="red")
abline(v=v[2], lwd=3, col="snow3")
abline(v=v[1], lwd=3, col="snow3")
abline(v=v[3], lwd=3, col="snow3")
abline(v=v[4], lwd=3, col="snow3")
```
**Comment** \newline
***A positive HR indicates worse conditions/prognosis, while a negative coefficient indicates a better condition/prognosis.***\newline
Riskscores > 1 corresponds to increased hazards of mortality with multiple HRF.
Riskscores < 1 corresponds to decreased hazards of mortality with multiple HRF, thus a survival benefit from chemotherapy.\newline 
The threshold for survival benefit is experienced when the risk score is < 1  at
which point the hazards of mortality is decreased.

\newpage
```{r}
plot(riskscore1_all, 1-lphr3, ylab = "Hazard Ratio", xlab="Risk Score",
     main= "Hazard Ratio vs RiskScores",col="blue")# ylim=c(0.6,1.6), xlim=c(0.6,1.6)

abline(h=1, lty=2, lwd=2, col="red")
abline(v=1, lwd=3, col="red")
abline(v=v[2], lwd=3, col="snow3")
abline(v=v[1], lwd=3, col="snow3")
abline(v=v[3], lwd=3, col="snow3")
abline(v=v[4], lwd=3, col="snow3")
```
***Subtracting HR from 1 gives reverse scale of HR reads, where lower HR indicates worse conditions, higher HR indicates better prognosis***\newline
As risk score increases, the hazard ratio gets worse (bad prognosis).
A survival benefit from adjuvant chemotherapy is realised when the risk score is < 1  at which point the hazard ratio of mortality appears to be better.




*****************************************************************
\newpage
\section{Riskscores from Est.PH}
# Est.PH {survC1} - Derivation of a risk score by a Cox proportioal hazard model
## Obtain Risk scores from the best predictors of mortality
```{r}
#Provides risk score by fitting data to a Cox's proportional hazards model with a given set of predictors
# Input data. The 1st column should be time-to-event, and the 2nd column is event indicator (1=event, 0=censor). The rest of the columns are covariates/predictors used in the model. No character variable or missing is allowed. 
#OUTPUT
# beta = Estimates for regression coefficient in the Cox model
# var = Variance-Covariance matrix for the beta above
# rs 	= Risk score of each individual
# ft 	= coxph object with the fitted model

library(survC1)
train1=lung[,c(1:2)] #time & status

train2 =lung[, c(3:15)] # other covariates

#convert other sub levels in all categorical covariates to integer
p = data.frame(lapply(train2, as.integer))

#combine numeric time & status with the numeric covariates
train_data = data.frame(cbind(train1,p)) 

#Make sure distribution of variables are not distorted after conversion
require(inspectdf)
show_plot(inspect_cat(train)) # inspect categorical columns 

show_plot(inspect_num(train)) #inspect numeric columns 

show_plot(inspect_num(train_data)) #inspect numeric columns 
```

```{r}
#obtain risk scores for each individual
rsmodel=Est.PH(train_data)
riskscores=rsmodel$rs
hist(riskscores)
hist(1-riskscores)
#riskscores1 = 1 - riskscores
```


```{r}
#obtain hazard rates .... didnt work with this approach
coef=rsmodel$beta
exp(coef)
hist(sqrt(riskscores))
#riskscores1 = 1 - riskscores
```


```{r}
# lphr3=predict(cox_fit3, lung, type="lp") #predicted hazard ratio
# hrrr=1-lphr3
```


```{r}
plot(riskscores, 1-lphr3, ylab = "Hazard Ratio" ,
     xlab="Risk Score", col="blue")# ylim=c(0.6,1.6), xlim=c(0.6,1.6)

abline(h=1, lty=2, lwd=2, col="red")
abline(v=1, lwd=3, col="red")
abline(v=v[2], lwd=3, col="snow3")
abline(v=v[1], lwd=3, col="snow3")
abline(v=v[3], lwd=3, col="snow3")
abline(v=v[4], lwd=3, col="snow3")
```















